---
title: "Dimensionality Reduction and Clustering for Palmer Penguins Dataset (Task 3)"
author: "Christian Rodriguez"
date: "2024-11-27"
output: 
  word_document:
    toc: true
    toc_depth: 3
---

# Introduction

This analysis focuses on dimensionality reduction and clustering techniques applied to the Palmer Penguins dataset. We perform Principal Component Analysis (PCA) to understand the variance structure and apply K-means clustering to identify natural groupings within the data, comparing them with actual species labels.

# Load Necessary Libraries

```{r setup, include=TRUE}
# Load required libraries
library(palmerpenguins)  # For the penguins dataset
library(dplyr)           # For data manipulation
library(ggplot2)         # For data visualization
library(factoextra)      # For PCA and clustering visualization
library(cluster)         # For clustering algorithms
```

# Load and Preprocess the Dataset

```{r load-data}
# Load the penguins dataset
data("penguins")
df <- penguins

# Impute missing numerical values with median by species
numerical_vars <- c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g")
df <- df %>%
  group_by(species) %>%
  mutate(across(all_of(numerical_vars), ~ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  ungroup()

# Function to calculate mode
get_mode <- function(v) {
  uniqv <- unique(v[!is.na(v)])
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

# Impute missing categorical values with mode by species
df <- df %>%
  group_by(species) %>%
  mutate(sex = ifelse(is.na(sex), get_mode(sex), sex)) %>%
  ungroup()

# Verify no missing values remain
colSums(is.na(df))
```

# Create Derived Variables

```{r create-bmi}
# Create a BMI-like variable
df <- df %>%
  mutate(
    weight_kg = body_mass_g / 1000,
    flipper_length_m = flipper_length_mm / 1000,
    BMI = weight_kg / (flipper_length_m^2)
  )
```

# Principal Component Analysis (PCA)

## Prepare Numerical Data for PCA

```{r prepare-pca}
# Select numerical variables for PCA
numerical_data <- df %>%
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g, BMI) %>%
  na.omit()

# Standardize the data
numerical_data_scaled <- scale(numerical_data)
```

## Perform PCA

```{r perform-pca}
# Perform PCA
pca_res <- prcomp(numerical_data_scaled, center = TRUE, scale. = TRUE)

# Summary of PCA results
summary(pca_res)
```

## Scree Plot to Determine Number of Components

```{r scree-plot, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
# Scree plot to visualize explained variance
fviz_eig(pca_res, addlabels = TRUE, ylim = c(0, 100)) +
  labs(title = "Scree Plot", x = "Principal Components", y = "Percentage of Variance Explained") +
  theme_minimal()
```

### Interpretation:

- **PC1:** Explains the highest variance, primarily influenced by variables like flipper length and body mass.
- **PC2:** Accounts for the second highest variance, driven by bill measurements.
- **Cumulative Variance:** The first two principal components explain a significant portion of the total variance (e.g., 80%), suggesting they capture the most critical information.

# PCA Biplot with Species Labels

```{r pca-biplot, fig.height=8, fig.width=10, warning=FALSE, message=FALSE}
# Biplot to visualize PCA results with species distinction
fviz_pca_biplot(pca_res, 
                geom = "point",
                habillage = df$species, 
                addEllipses = TRUE, 
                ellipse.level = 0.95,
                palette = "jco",
                repel = TRUE) +
  labs(title = "PCA Biplot of Palmer Penguins") +
  theme_minimal()
```

### Interpretation:

- **Species Separation:** Different penguin species cluster distinctly in the PCA space, indicating that the principal components effectively capture species-specific traits.
- **Variable Contributions:** Arrows in the biplot show the direction and magnitude of variable contributions to each principal component.

# K-Means Clustering

## Determine Optimal Number of Clusters Using the Elbow Method

```{r elbow-method, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
# Determine the optimal number of clusters using the Elbow method
set.seed(123)
fviz_nbclust(numerical_data_scaled, kmeans, method = "wss") +
  labs(title = "Elbow Method for Determining Optimal Clusters") +
  theme_minimal()
```

### Interpretation:

- **Elbow Point:** The plot suggests that 3 clusters are optimal, corresponding to the three penguin species.

## Apply K-Means Clustering with 3 Clusters

```{r kmeans-clustering}
# Apply K-means clustering with 3 centers
set.seed(123)
kmeans_res <- kmeans(numerical_data_scaled, centers = 3, nstart = 25)

# Add cluster assignments to the dataset
numerical_data$cluster <- as.factor(kmeans_res$cluster)
```

## Visualize Clusters

```{r visualize-clusters, fig.height=8, fig.width=10, warning=FALSE, message=FALSE}
# Visualize the clustering results in PCA space
fviz_cluster(kmeans_res, 
             data = numerical_data_scaled,
             geom = "point",
             ellipse.type = "norm",
             palette = "jco",
             ggtheme = theme_minimal(),
             main = "K-Means Clustering with 3 Clusters") +
  labs(color = "Cluster")
```

## Compare Clusters with Actual Species Labels

```{r cluster-species-comparison}
# Create a table to compare clusters with actual species
cluster_species_table <- table(kmeans_res$cluster, df$species[complete.cases(df)])
print(cluster_species_table)
```

### Interpretation:

- **High Agreement:** The majority of penguins are correctly clustered according to their species, validating the effectiveness of K-means clustering on this dataset.
- **Misclassifications:** Any discrepancies can provide insights into overlapping characteristics or potential data issues.

# Conclusion

The PCA effectively reduced the dataset's dimensionality while retaining a significant portion of the variance, facilitating visualization and interpretation. K-means clustering successfully identified clusters that closely align with the actual penguin species, demonstrating clear differentiation based on morphological measurements.

# Recommendations

- **Advanced Clustering Techniques:** Explore hierarchical clustering or DBSCAN for potentially better cluster identification.
- **Feature Engineering:** Incorporate additional derived variables to enhance clustering performance.
- **Validation:** Use external validation metrics or cross-validation to further assess clustering robustness.
