---
title: "Data Preprocessing for Palmer Penguins Dataset (Task 1)"
author: "Christian Rodriguez"
date: "2024-11-27"
output: 
  word_document: 
    toc: true
    toc_depth: 3
---

# Introduction

In this analysis, we perform data preprocessing on the Palmer Penguins dataset. The steps include loading and inspecting the dataset, handling missing values through logical imputation, and conducting an extensive exploratory data analysis (EDA) with summary statistics and visualizations.

# Load Necessary Libraries

```{r setup, include=TRUE}
# Load required libraries
library(palmerpenguins)  # For the penguins dataset
library(dplyr)           # For data manipulation
library(ggplot2)         # For data visualization
library(tidyr)           # For data tidying
library(GGally)          # For pair plots
```

# Load the Dataset

```{r load-data}
# Load the penguins dataset
data("penguins")
df <- penguins
```

# Inspect the Dataset

```{r inspect-data}
# Structure of the dataset
str(df)

# Summary statistics
summary(df)

# Preview the first few rows
head(df)
```

# Handle Missing Values

## Check for Missing Values

```{r missing-values}
# Calculate the number of missing values in each column
colSums(is.na(df))
```

## Impute Missing Numerical Values

### Strategy

Imputing missing numerical values using the median within each species ensures that the imputed values are representative of the specific traits of that species. This approach maintains the integrity of species-specific distributions and avoids introducing bias that could occur if a global median were used.

```{r impute-numerical}
# Define numerical variables
numerical_vars <- c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g")

# Impute missing numerical values with median by species
df <- df %>%
  group_by(species) %>%
  mutate(across(all_of(numerical_vars), ~ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  ungroup()
```

## Impute Missing Categorical Values

### Strategy

For categorical variables, imputing missing values with the mode within each species helps maintain the natural distribution and relationships in the data. This method ensures that the imputed categories are consistent with the species-specific characteristics.

```{r impute-categorical}
# Function to calculate mode
get_mode <- function(v) {
  uniqv <- unique(v[!is.na(v)])
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

# Impute missing values in 'sex' variable with mode by species
df <- df %>%
  group_by(species) %>%
  mutate(sex = ifelse(is.na(sex), get_mode(sex), sex)) %>%
  ungroup()
```

## Verify No Missing Values Remain

```{r verify-no-missing}
# Confirm that there are no missing values
colSums(is.na(df))
```

# Create Derived Variables

### Example: Creating a BMI-like Variable

Assuming `body_mass_g` as weight in grams and `flipper_length_mm` as height in millimeters, we create a BMI-like metric for penguins.

```{r create-bmi}
# Create a new column 'BMI' based on 'body_mass_g' and 'flipper_length_mm'
# Convert grams to kilograms and millimeters to meters for BMI calculation
df <- df %>%
  mutate(
    weight_kg = body_mass_g / 1000,
    flipper_length_m = flipper_length_mm / 1000,
    BMI = weight_kg / (flipper_length_m^2)
  )

# View the first few rows to verify the new columns
head(df)
```

# Generate Summary Statistics

```{r summary-statistics}
# Summary statistics of the cleaned dataset
summary(df)
```

# Initial Visualizations

## Boxplots for Numerical Variables by Species

```{r boxplots, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
# Boxplots to visualize the distribution of numerical variables across species
ggplot(df, aes(x = species, y = bill_length_mm, fill = species)) +
  geom_boxplot() +
  labs(title = "Bill Length by Species", x = "Species", y = "Bill Length (mm)") +
  theme_minimal()

ggplot(df, aes(x = species, y = bill_depth_mm, fill = species)) +
  geom_boxplot() +
  labs(title = "Bill Depth by Species", x = "Species", y = "Bill Depth (mm)") +
  theme_minimal()

ggplot(df, aes(x = species, y = flipper_length_mm, fill = species)) +
  geom_boxplot() +
  labs(title = "Flipper Length by Species", x = "Species", y = "Flipper Length (mm)") +
  theme_minimal()

ggplot(df, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot() +
  labs(title = "Body Mass by Species", x = "Species", y = "Body Mass (g)") +
  theme_minimal()
```

## Scatterplots to Explore Relationships Between Variables

```{r scatterplots, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
# Scatterplots to identify relationships between variables colored by species
ggplot(df, aes(x = bill_length_mm, y = body_mass_g, color = species)) +
  geom_point(alpha = 0.7) +
  labs(title = "Body Mass vs. Bill Length", x = "Bill Length (mm)", y = "Body Mass (g)") +
  theme_minimal()

ggplot(df, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(alpha = 0.7) +
  labs(title = "Body Mass vs. Flipper Length", x = "Flipper Length (mm)", y = "Body Mass (g)") +
  theme_minimal()

ggplot(df, aes(x = bill_depth_mm, y = flipper_length_mm, color = species)) +
  geom_point(alpha = 0.7) +
  labs(title = "Flipper Length vs. Bill Depth", x = "Bill Depth (mm)", y = "Flipper Length (mm)") +
  theme_minimal()
```

## Additional EDA

### Histograms by Species

```{r histograms, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
# Histograms to visualize the distribution of body mass by species
ggplot(df, aes(x = body_mass_g, fill = species)) +
  geom_histogram(binwidth = 200, alpha = 0.7, position = "identity") +
  facet_wrap(~ species) +
  labs(title = "Distribution of Body Mass by Species", x = "Body Mass (g)", y = "Count") +
  theme_minimal()
```

### Pair Plot

```{r pair-plot, fig.height=8, fig.width=10, warning=FALSE, message=FALSE}
# Pairwise scatter plots with correlations
ggpairs(df, columns = c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g", "BMI"),
        aes(color = species, alpha = 0.5)) +
  theme_minimal()
```

# Interpretation

-   **Boxplots:** Reveal distinct differences in measurements among species, indicating unique physical characteristics.
    -   **Distribution of Body Mass by Species:**\
        The histogram clearly illustrates the distribution of body mass for each penguin species:
        -   **Adelie Penguins (Red):** Show a relatively narrow range of body mass, clustering around 3,500–4,000 g, indicating a smaller size compared to other species.

        -   **Chinstrap Penguins (Green):** Display a slightly broader range, primarily concentrated between 3,500–4,200 g, with some overlap with Adelie penguins.

        -   **Gentoo Penguins (Blue):** Exhibit a distinct and much larger body mass distribution, centered around 5,000–6,000 g, with no overlap with the other two species, highlighting their significantly larger size.
-   **Histograms:**\
    The histograms reveal distinct distributions of morphological measurements (e.g., bill length, bill depth, flipper length, and body mass) across the three penguin species. Each species exhibits unique patterns, with minimal overlap in some traits, such as flipper length for Gentoo penguins.

<!-- -->

-   **Scatterplots:**\
    The scatterplots highlight relationships between variables, such as the positive correlation between flipper length and body mass. Clear clusters corresponding to each penguin species are observed, particularly for Gentoo penguins, which stand out due to their larger body mass and flipper length.

-   **Pair Plots:**\
    The pair plots provide insights into variable interactions and species-specific trends:

    -   **Adelie Penguins (Red):** Generally smaller in size with less variation in morphological traits.

    -   **Chinstrap Penguins (Green):** Show intermediate values with noticeable overlap in some traits with Adelie penguins.

    -   **Gentoo Penguins (Blue):** Distinctly larger in size, standing out in flipper length and body mass.

-   **Correlations:**\
    The correlation coefficients across variables differ by species:

    -   High positive correlations are observed for Gentoo penguins between flipper length and body mass.

    -   Chinstrap penguins exhibit stronger correlations in bill depth and body mass compared to other species.

# Conclusion

The data preprocessing steps effectively cleaned the Palmer Penguins dataset by addressing missing values and generating derived variables where necessary. The exploratory data analysis highlights clear distinctions among penguin species based on key morphological measurements, such as body mass, flipper length, and bill dimensions. These analyses provide valuable insights into species-specific traits and relationships between variables, establishing a solid foundation for further statistical analysis, clustering, and predictive modeling in subsequent tasks.
