---
title: "Dimensionality Reduction and Clustering for Palmer Penguins Dataset (Task 3)"
author: "Christian Rodriguez"
date: "2024-11-27"
output: 
  word_document:
    toc: true
    toc_depth: 3
---

# Introduction

This analysis focuses on dimensionality reduction and clustering techniques applied to the Palmer Penguins dataset. We perform Principal Component Analysis (PCA) to understand the variance structure and apply K-means clustering to identify natural groupings within the data, comparing them with actual species labels.

# Load Necessary Libraries

```{r setup, include=TRUE}
# Load required libraries
library(palmerpenguins)  # For the penguins dataset
library(dplyr)           # For data manipulation
library(ggplot2)         # For data visualization
library(factoextra)      # For PCA and clustering visualization
library(cluster)         # For clustering algorithms
```

# Load and Preprocess the Dataset

```{r load-data}
# Load the penguins dataset
data("penguins")
df <- penguins

# Impute missing numerical values with median by species
numerical_vars <- c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g")
df <- df %>%
  group_by(species) %>%
  mutate(across(all_of(numerical_vars), ~ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  ungroup()

# Function to calculate mode
get_mode <- function(v) {
  uniqv <- unique(v[!is.na(v)])
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

# Impute missing categorical values with mode by species
df <- df %>%
  group_by(species) %>%
  mutate(sex = ifelse(is.na(sex), get_mode(sex), sex)) %>%
  ungroup()

# Verify no missing values remain
colSums(is.na(df))
```

# Create Derived Variables

```{r create-bmi}
# Create a BMI-like variable
df <- df %>%
  mutate(
    weight_kg = body_mass_g / 1000,
    flipper_length_m = flipper_length_mm / 1000,
    BMI = weight_kg / (flipper_length_m^2)
  )
```

# Principal Component Analysis (PCA)

## Prepare Numerical Data for PCA

```{r prepare-pca}
# Select numerical variables for PCA
numerical_data <- df %>%
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g, BMI) %>%
  na.omit()

# Standardize the data
numerical_data_scaled <- scale(numerical_data)
```

## Perform PCA

```{r perform-pca}
# Perform PCA
pca_res <- prcomp(numerical_data_scaled, center = TRUE, scale. = TRUE)

# Summary of PCA results
summary(pca_res)
```

## Scree Plot to Determine Number of Components

```{r scree-plot, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
# Scree plot to visualize explained variance
fviz_eig(pca_res, addlabels = TRUE, ylim = c(0, 100)) +
  labs(title = "Scree Plot", x = "Principal Components", y = "Percentage of Variance Explained") +
  theme_minimal()
```

### Interpretation:

-   **PC1 (59.1% Variance Explained):**\
    The first principal component captures the largest portion of variance in the dataset (59.1%), primarily influenced by variables like **flipper length** and **body mass**, which are strongly correlated with overall penguin size and weight.

<!-- -->

-   **PC2 (21% Variance Explained):**\
    The second principal component accounts for an additional 21% of the variance, mainly driven by **bill measurements** (bill length and depth). This component likely captures variation related to penguin species and dietary adaptations.

-   **Cumulative Variance:**\
    The first two principal components together explain **80.1% of the total variance**, indicating they capture most of the critical information in the dataset. This suggests that these two components are sufficient for dimensionality reduction while retaining essential variability in the data.

-   **PC3 to PC5 (Remaining Variance):**\
    The remaining components (PC3, PC4, and PC5) contribute progressively less to the explained variance, with PC5 contributing virtually nothing. These components are less significant and likely capture noise or very specific, minor variations in the data.

# PCA Biplot with Species Labels

```{r pca-biplot, fig.height=8, fig.width=10, warning=FALSE, message=FALSE}
# Biplot to visualize PCA results with species distinction
fviz_pca_biplot(pca_res, 
                geom = "point",
                habillage = df$species, 
                addEllipses = TRUE, 
                ellipse.level = 0.95,
                palette = "jco",
                repel = TRUE) +
  labs(title = "PCA Biplot of Palmer Penguins") +
  theme_minimal()
```

### Interpretation:

-   <div>

    #### **Species Separation:**

    </div>

    -   The PCA biplot shows clear clustering of the three penguin species (**Adelie**, **Chinstrap**, and **Gentoo**) in the PCA space, suggesting that the principal components effectively distinguish species-specific traits.

        -   **Adelie Penguins (Blue):** Cluster near the center-left, characterized by smaller body mass and shorter flipper lengths.

        -   **Chinstrap Penguins (Yellow):** Overlap slightly with Adelie but are positioned closer to the center, reflecting intermediate trait values.

        -   **Gentoo Penguins (Gray):** Form a distinct cluster to the far right, indicating larger body mass and longer flipper lengths.

    #### **Variable Contributions:**

    -   The arrows represent the contributions of variables to the principal components:

        -   **Flipper Length (Dim1):** Strongly aligned with Dim1, indicating that it contributes significantly to the first principal component and is positively correlated with body mass.

        -   **Body Mass (Dim1):** Similarly contributes heavily to Dim1, highlighting its importance in explaining the variance in penguin size.

        -   **Bill Length and Bill Depth (Dim2):** These variables contribute more to Dim2, indicating they capture species-specific differences not fully explained by body size.

        -   **BMI (Dim2):** Points in a different direction, suggesting it explains variance orthogonal to size-related traits.

    #### **Principal Component Contributions:**

    -   **Dim1 (59.1% Variance Explained):** Primarily explains variation in penguin size, driven by flipper length and body mass.

    -   **Dim2 (21% Variance Explained):** Captures differences related to bill characteristics, helping separate species with similar body sizes.

    ### Summary:

    The PCA biplot effectively visualizes the separation of penguin species based on their morphological traits. **Gentoo Penguins** are distinct due to their larger size, while **Adelie** and **Chinstrap Penguins** show more overlap, with slight distinctions driven by bill characteristics. The contributions of flipper length, body mass, and bill dimensions to the principal components align well with biological expectations.

# K-Means Clustering

## Determine Optimal Number of Clusters Using the Elbow Method

```{r elbow-method, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
# Determine the optimal number of clusters using the Elbow method
set.seed(123)
fviz_nbclust(numerical_data_scaled, kmeans, method = "wss") +
  labs(title = "Elbow Method for Determining Optimal Clusters") +
  theme_minimal()
```

### Interpretation:

-   **Elbow Point:**\
    The "elbow" in the plot, where the total within-cluster sum of squares (WSS) starts to level off, occurs at **k = 3**. This suggests that dividing the data into 3 clusters provides an optimal balance between minimizing within-cluster variation and maintaining a manageable number of clusters.
-   **Relevance to Penguins Dataset:**\
    The identification of 3 optimal clusters aligns with the biological understanding of the dataset, where there are three distinct penguin species (**Adelie**, **Chinstrap**, and **Gentoo**). This supports the hypothesis that the clustering captures meaningful species-level differences in the data.

### Summary:

The Elbow Method confirms that **3 clusters** are optimal, consistent with the known species structure in the dataset. This result reinforces the appropriateness of clustering as a technique for analyzing the penguin dataset.

## Apply K-Means Clustering with 3 Clusters

```{r kmeans-clustering}
# Apply K-means clustering with 3 centers
set.seed(123)
kmeans_res <- kmeans(numerical_data_scaled, centers = 3, nstart = 25)

# Add cluster assignments to the dataset
numerical_data$cluster <- as.factor(kmeans_res$cluster)
```

## Visualize Clusters

```{r visualize-clusters, fig.height=8, fig.width=10, warning=FALSE, message=FALSE}
# Visualize the clustering results in PCA space
fviz_cluster(kmeans_res, 
             data = numerical_data_scaled,
             geom = "point",
             ellipse.type = "norm",
             palette = "jco",
             ggtheme = theme_minimal(),
             main = "K-Means Clustering with 3 Clusters") +
  labs(color = "Cluster")
```

## Compare Clusters with Actual Species Labels

```{r cluster-species-comparison}
# Create a table to compare clusters with actual species
cluster_species_table <- table(kmeans_res$cluster, df$species[complete.cases(df)])
print(cluster_species_table)
```

### Interpretation:

-   **High Agreement:**

    -   The clustering results from K-means show a high level of agreement with the actual species labels:

        -   **Cluster 1:** Contains 124 Gentoo penguins (perfect match with the Gentoo species).

        -   **Cluster 2:** Predominantly consists of 74 Adelie penguins and 58 Chinstrap penguins, showing overlap between these species.

        -   **Cluster 3:** Includes 78 Adelie penguins and 10 Chinstrap penguins, further indicating some misclassification between these two species.

-   **Misclassifications:**

    -   The overlap in **Cluster 2** and **Cluster 3** between Adelie and Chinstrap penguins suggests that these species share similar morphological traits, making them harder to distinguish through clustering.

    -   The absence of Gentoo penguins in Clusters 2 and 3 highlights their distinct traits, making them easier to separate from the other species.

-   **Biological Insights:**

    -   Misclassifications between Adelie and Chinstrap penguins might reflect shared habitats or intermediate body characteristics. Further exploration of specific variables, such as bill length and depth, could help clarify the overlap.

### Summary:

The K-means clustering results align well with the actual species labels, demonstrating the effectiveness of this method in classifying penguin species. While Gentoo penguins are perfectly separated, the overlap between Adelie and Chinstrap species highlights areas for further analysis, potentially refining the clustering process or identifying overlapping features in these groups.

# Conclusion

The PCA effectively reduced the dataset's dimensionality, retaining a substantial portion of the variance (80.1% in the first two components), and enabled clear visualization of species-level differences. K-means clustering complemented this by successfully identifying clusters that closely align with the actual penguin species. While Gentoo penguins were distinctly separated, some overlap between Adelie and Chinstrap penguins was observed, reflecting their shared morphological traits. Overall, the combination of PCA and K-means clustering provided meaningful insights into species differentiation based on morphological measurements.
